---
title: "Default priors"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Default priors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = F}
library(knitr)
knitr::opts_chunk$set(
  out.extra = 'style="display:block; margin:auto;"'
)
```

A LogNormal prior is placed on $R$, and a HalfNormal on $1 / \sqrt{k}$.
The default prior on $R$ stipulates that $R$ is most likely between 1/2 and 2, while the default prior on $k$ puts equal amounts of mass in both the strongly-overdispersed $k < 1$ regime and the less-overdispersed $k \geq 1$ regime.
Samples from this are available in the package data object `prior_predictive`, which we will use to visualize the priors.

## Default prior on $R$
```{r, echo=FALSE, message=FALSE}
library(nbbp)
library(ggplot2)
data(prior_predictive)

prior_predictive |>
  ggplot(aes(x = r_eff, y = after_stat(density))) +
  geom_histogram() +
  theme_minimal() +
  xlab("R")
```

## Default prior on $k$

The (implicit) default prior on $k$ has heavy tails, reflecting the prior's pull towards the Poisson regime of $k \to \infty$.
Truncating to $k \leq 100$ for visualization purposes, the (conditional) prior is
```{r, echo=FALSE, message=FALSE}
prior_predictive |>
  dplyr::filter(dispersion <= 10) |>
  ggplot(aes(x = dispersion, y = after_stat(density))) +
  geom_histogram() +
  theme_minimal() +
  xlab("k")
```

## Default prior predictive offspring distribution
```{r, echo=FALSE, message=FALSE}
prior_predictive |>
  ggplot(aes(x = offspring_count, y = after_stat(density))) +
  geom_histogram(binwidth = 1) +
  theme_minimal() +
  xlab("# offspring")
```

## Default prior predictive chain size distribution

The prior predictive distribution on chain sizes has two components.
There is a `r {round(mean(is.infinite(prior_predictive$chain_size)), 2) * 100}`% chance of a non-extinct (infinite) chain (see the vignette "Advanced data" for more on this).
Conditional on extinction (finite size), there are long tails, with a `r round(mean(prior_predictive$chain_size > 100), 2) * 100`% chance of a chain size above 100.
Focusing in on chains of 100 or fewer for visualization, the (conditional) prior distribution is:
```{r, echo=FALSE, message=FALSE}
prior_predictive |>
  dplyr::filter(chain_size <= 100) |>
  ggplot(aes(x = chain_size, y = after_stat(density))) +
  geom_histogram(binwidth = 1) +
  theme_minimal() +
  xlab("chain size")
```
